---
title: "DO data types"
output: 
    html_notebook:
        toc: true
        toc_float: true
        code_folding: hide
        df_print: paged
---

# Purpose

To identify the most appropriate format for various data in the Human Disease Ontology. _For information regarding definitions, see the corresponding R notebook._

```{r setup, include = FALSE}
library(tidyverse)
library(DO.utils)
```

```{r init_repo}
r <- DO.utils::DOrepo("~/Documents/Ontologies/HumanDiseaseOntology")
```


# Synonyms

```{r}
q <- '# Return a list of all non-obsolete synonyms with metadata

PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
PREFIX owl: <http://www.w3.org/2002/07/owl#>
PREFIX oboInOwl: <http://www.geneontology.org/formats/oboInOwl#>

SELECT ?id ?label ?syn_type ?synonym ?lang ?data_type
WHERE {
    ?class a owl:Class ;
        ?pred ?synonym .
    FILTER(CONTAINS(STR(?class), "DOID"))
    FILTER(REGEX(STR(?pred), "[Ss]ynonym"))
    
	OPTIONAL { ?class oboInOwl:id ?id . }
    OPTIONAL { ?class rdfs:label ?label . }
    
    BIND(STRAFTER(STR(?pred), "oboInOwl#") AS ?syn_type)
    BIND(LANG(?synonym) AS ?lang)
    BIND(DATATYPE(?synonym) AS ?data_type)
    FILTER NOT EXISTS { ?class owl:deprecated ?obsolete . }
}'

syn <- r$doid$query(q) %>%
    tibble::as_tibble()
```


1. Are synonyms commonly formatted with a language tag (`@en`) or RDF data type (`xsd:string`)?

Only one of language tag/data type can be specified for a literal but language tag always implies the string data type. The ratio of synonyms with language tags is then the indicator of which is more common.

```{r}
dplyr::count(syn, lang)
dplyr::count(syn, data_type)
```

**Most synonyms have the English language tag (`@en`).**


2. What synonyms types are included in DO and how common is each?

```{r}
dplyr::count(syn, syn_type, sort = TRUE)
```

Most are exact but a number of types are used.


# Xref

```{r}
q <- '# Return a list of all non-obsolete cross-references with metadata

PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
PREFIX owl: <http://www.w3.org/2002/07/owl#>
PREFIX oboInOwl: <http://www.geneontology.org/formats/oboInOwl#>

SELECT ?id ?label ?xref ?lang ?data_type
WHERE {
    ?class a owl:Class ;
        oboInOwl:hasDbXref ?xref .
    FILTER(CONTAINS(STR(?class), "DOID"))
    
	OPTIONAL { ?class oboInOwl:id ?id . }
    OPTIONAL { ?class rdfs:label ?label . }
    
    BIND(LANG(?synonym) AS ?lang)
    BIND(DATATYPE(?synonym) AS ?data_type)
    FILTER NOT EXISTS { ?class owl:deprecated ?obsolete . }
}'

xref <- r$doid$query(q) %>%
    tibble::as_tibble() %>%
    tidyr::unnest(lang, keep_empty = TRUE) %>%
    tidyr::unnest(data_type, keep_empty = TRUE)
```


1. Are xrefs commonly formatted with a language tag (`@en`) or RDF data type (`xsd:string`)?

```{r}
dplyr::count(xref, lang)
dplyr::count(xref, data_type)
```

Neither of those exist in the doid.owl file. This might be due to the use of ROBOT to create doid.owl.


2. For the sake of editing purposes, how are these handled in the doid-edit.owl file?

```{r}
de <- DO.utils:::read_doid_edit("~/Documents/Ontologies/HumanDiseaseOntology")
de_xref <- tibble(
    raw = de[
        stringr::str_detect(de, "hasDbXref") & !stringr::str_detect(de,"IAO") &
            stringr::str_detect(de, "DOID")
    ],
    data_type = dplyr::case_when(
        stringr::str_detect(raw, "@en") ~ "lang:en",
        stringr::str_detect(raw, "xsd:string") ~ "xsd:string",
        TRUE ~ NA_character_
    )
)
```

```{r}
dplyr::count(de_xref, data_type)
```

Overwhelmingly, use `xsd:string`.
