---
title: "Understanding DO's Imports"
output: 
    html_notebook:
        toc: true
        toc_float: true
        df_print: paged
---

# Background

Currently all indirect imports are processed with a common `robot remove` command when created by the Makefile in src/ontology/imports/:

```
remove \
	 --select complement \
	 --select annotation-properties \
	 --term-file annotations.txt \
```

While annotations.txt in src/ontology/imports contains 4 elements:

- rdfs:label
- oboInOwl:id
- obo:IAO_0000115
- owl:deprecated

This command is designed to retain only these 4 elements (it removes their complement = all axioms that includes any annotation property other than these 4 elements).


## Con

The downside to this approach is that all definitions with oboInOwl:hasDbXref source annotations result in the entire definition being removed.


```{r setup, include = FALSE}
library(tidyverse)
library(DO.utils)
```


```{r}
# list imports
import_files <- list.files("/Users/allenbaron/Documents/Ontologies/HumanDiseaseOntology/src/ontology/imports/build", pattern = "^[a-z]+\\.owl")

# setup imports for queries
imports <- purrr::map(import_files, owl_xml)
```


```{r}
# explore annotated{.} in 
```

