---
title: "Anaylsis: DO rare disease subset (DO_rare_slim)"
author: "J. Allen Baron"
output: 
    html_notebook:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---


# Purpose

To determine how rare diseases are currently organized in DO and identify
whether manual curation is needed.

More specifically, the following questions will be addressed:

1. Are there diseases in the rare disease subset _without_ a GARD or ORDO xref?
    - If so, how many and what are they?
2. Are there diseases with a GARD or ORDO xref that are not in the rare disease subset?
    - If so, how many and what are they?


# Data Preparation

Diseases with their xrefs will be exported from doid.owl (full disease set) and DO_rare_slim.owl (rare disease subset).
```{r setup, include = FALSE}
library(here)
library(DO.utils)
library(tidyverse)
library(googledrive)
library(googlesheets4)
```

```{r declare_io, message = FALSE}
# input
do_repo <- here::here("../Ontologies/HumanDiseaseOntology")
do_owl <- file.path(do_repo, "src/ontology/doid.owl")
rare_owl <- file.path(do_repo, "src/ontology/subsets/DO_rare_slim.owl")

# intermediate
disease_tsv <- here::here("data/DO/all_disease-xref.tsv")
rare_dis_tsv <- here::here("data/DO/rare_disease-xref.tsv")

# output
rare_out <- c(no_xref = "in_set-no_xref", not_in = "xref-not_in_set")
curate_gs <- googlesheets4::gs4_create(
    "Curate-Rare_subset",
    sheets = unname(rare_out)
)
googledrive::drive_mv(curate_gs, path = "My_Work/Curation/", overwrite = FALSE)
```

```{r robot_export}
# export xrefs from full ontology and rare disease subset for comparison
robot_export <- function(input, output) {
    system2(
        'robot',
        glue::glue(
            'export --input {input} ',
            '--header="ID|LABEL|oboInOwl:hasDbXref" ',
            '--export {output}',
            input = input,
            output = output
        )
    )
}

do_status <- robot_export(do_owl, disease_tsv)
rare_status <- robot_export(rare_owl, rare_dis_tsv)


if (do_status == 0 && rare_status == 0) {
    message("Diseases and xrefs exported successfully!")
}
```

```{r load_data, message=FALSE}
disease <- readr::read_tsv(disease_tsv) %>%
    dplyr::rename(id = ID, label = LABEL, xref = 'oboInOwl:hasDbXref') %>%
    dplyr::filter(
        stringr::str_detect(id, "DOID"),
        !stringr::str_detect(label, "obsolete")
    )
rare <- readr::read_tsv(rare_dis_tsv) %>%
    dplyr::rename(id = ID, label = LABEL, xref = 'oboInOwl:hasDbXref')
```


# 1. Diseases in Subset _without_ GARD or ORDO xref

```{r rare_no_xref}
rare_no_xref <- dplyr::filter(rare, !stringr::str_detect(xref, "GARD|ORDO"))
```

There are `r nrow(rare_no_xref)` diseases in DO_rare_slim that do not have a GARD or ORDO xref. These should be reviewed.

```{r show_rnx, echo = FALSE}
googlesheets4::write_sheet(
    rare_no_xref,
    ss = curate_gs,
    sheet = unname(rare_out['no_xref'])
)
rare_no_xref
```


# 2. Missing Diseases with GARD or ORDO xref

```{r fn-extract_concat, include = FALSE}
extract_concat <- function(x, pattern) {
    purrr::map_chr(
        x,
        ~ stringr::str_extract_all(.x, pattern) %>%
            unlist() %>%
            DO.utils::vctr_to_string() %>%
            DO.utils::replace_blank()
    )
}
```


```{r rare_missing}
rare_missing <- dplyr::anti_join(disease, rare, by = "id") %>%
    dplyr::filter(stringr::str_detect(xref, "GARD|ORDO")) %>%
    dplyr::mutate(
        gard = extract_concat(xref, "GARD:[0-9]+"),
        ordo = extract_concat(xref, "ORDO:[0-9]+")
    ) %>%
    dplyr::select(-xref)
```

There are `r format(nrow(rare_missing), big.mark = ",")` diseases with ORDO or GARD xrefs that are not in the DO_rare_slim. These should be added.

```{r show_missing, echo = FALSE}
googlesheets4::write_sheet(
    rare_missing,
    ss = curate_gs,
    sheet = unname(rare_out['not_in'])
)
rare_missing
```


# Summary

- Total Diseases: `r format(nrow(disease), big.mark = ",")`
    - In DO_rare_slim: `r format(nrow(rare), big.mark = ",")`
    - With rare xref, not in slim: `r format(nrow(rare_missing), big.mark = ",")`
    - Non-rare: `r format(nrow(dplyr::filter(disease, !id %in% rare_missing$id, !id %in% rare$id)), big.mark = ",")`
    - Possible non-rare: `r format(nrow(rare_no_xref), big.mark = ",")` _(in DO_rare_slim without rare xref)_
