---
title: "Comparison: DO and GARD"
date: "2023-01-18"
output:
    html_notebook:
        toc: true
        toc_float: true
        code_folding: hide
        df_print: paged
        number_sections: true
---


# Purpose

To determine how many [GARD](https://rarediseases.info.nih.gov) diseases are in the Human Disease Ontology.

```{r setup, include = FALSE}
library(tidyverse)
library(DO.utils)
library(janitor)
library(here)
library(git2r)
library(lubridate)
library(googlesheets4)

# curation google sheet
gard_gs <- "183ZmU9Y2JwyUkw1AgaH-hDrDSYOXH5BiAj_IPwOcKgk"
```


# Data

## GARD

All GARD diseases are extracted from the GARD sitemap. Data include GARD numeric identifiers and names (with some slight formatting differences relative to presented names).
```{r gard_csv_data, message = FALSE}
gard_csv <- here::here("data/mapping/GARD Disease List Website_Feb2023.csv")
gard_sourced <- readr::read_csv(gard_csv) %>%
    janitor::clean_names() %>%
    dplyr::mutate(
        label_std = source_name %>%
            stringr::str_to_lower() %>%
            stringr::str_remove_all(
                "([:blank:]|[:punct:]|[:space:])+"
            )
    )
```


```{r gard_sitemap_data, message = FALSE}
gard_sitemap_tsv <- here::here("data/mapping/GARD.tsv")
if (file.exists(gard_sitemap_tsv)) {
    gard <- readr::read_tsv(gard_sitemap_tsv)
} else {
    gard_xml <- readr::read_file("https://rarediseases.info.nih.gov/sitemap.xml")
    
    gard_info <- stringr::str_match_all(
        gard_xml,
        "https://rarediseases.info.nih.gov/diseases/([0-9]+)/([^< /\"]+)"
    )[[1]]
    
    gard <- tibble::tibble(
        gard_id = paste0("GARD:", gard_info[, 2]),
        label = stringr::str_replace_all(gard_info[, 3], "-", " ")
    ) %>%
        unique()

    readr::write_tsv(gard, gard_sitemap_tsv)
}
```

GARD sitemap data include `r format(nrow(gard), big.mark = ",")` diseases, which matches the number of diseases listed on GARD's [browser](https://rarediseases.info.nih.gov/diseases) on the date this analysis started (2023-01-18).

The data in the downloaded csv has `r format(nrow(gard_sourced), big.mark = ",")` along with synonyms and Orphanet/OMIM xrefs. This dataset appears to be entirely derived from Orphanet and OMIM without original GARD data, which agrees with the approach Eric at NCATS indicated GARD is adopting (i.e. no original data in GARD).

The GARD labels used on the website do not all match those in the .csv file, so it makes sense to create a new section for comparison and a corresponding new curation document.

```{r}
dplyr::anti_join(
    dplyr::mutate(gard, label = stringr::str_remove_all(label, " ")),
    gard_sourced,
    by = c("label" = "label_std")
)
```

## Human Disease Ontology

```{r do_repo}
repo <- here::here("../Ontologies/HumanDiseaseOntology")
repo_tags <- tibble::tibble(
    tag = git2r::tags(repo),
    label = names(tag),
    date = lubridate::fast_strptime(
        label,
        format = c("v%Y-%m-%d", "v%Y-%m-%dv2", "%Y-%m-%d", "vv%Y-%m-%d",
                   "v%Y-%m0%d",  "v%Y-%m/%d", "v%Y-%m-%d.2", "v%Y-%m-%d_v2")
    )
)
latest_tag <- repo_tags %>%
    dplyr::filter(date == max(date)) %>%
    .$tag
```

DO data is from release `r names(latest_tag)`.

```{r query_fxn, include = FALSE}
query_release_doid <- function(repo, release, query) {
    head <- git2r::repository_head(repo)
    on.exit(git2r::checkout(head))
    
    git2r::checkout(release)

    do_owl <<- DO.utils::owl_xml(file.path(repo, "src/ontology/doid.owl"))

    do_owl$query(query) %>%
        DO.utils::tidy_sparql()
}
```


```{r do_data}
do <- query_release_doid(
    repo,
    latest_tag[[1]],
    query = "
        SELECT ?doid ?label ?gard_id
        WHERE {
            ?class a owl:Class ;
                oboInOwl:id ?doid ;
                rdfs:label ?label .
            OPTIONAL {
                ?class oboInOwl:hasDbXref ?gard_id .
                FILTER(CONTAINS(str(?gard_id), 'GARD'))
            }
            FILTER NOT EXISTS { ?class owl:deprecated ?any . }
        }"
)

do_syn <-  do_owl$query("
    SELECT ?doid ?synonym
    WHERE {
        ?class a owl:Class ;
            oboInOwl:id ?doid ;
            oboInOwl:hasExactSynonym ?synonym .

        FILTER NOT EXISTS { ?class owl:deprecated ?any . }
    }"
) %>%
    tidy_sparql()
```


## Mondo

Mondo is used only as a bridge to identify DO-GARD relationships via cross-references. Mondo 2023-01-04 release is used in this analysis.

```{r mondo_data}
mondo_owl <- DO.utils::owl_xml(
    here::here("../Ontologies/mondo/mondo-20230104.owl")
)

mondo_bridge <- mondo_owl$query(
    "SELECT ?doid ?mondo_gard_id ?mondo_label
    WHERE {
        ?class a owl:Class ;
            oboInOwl:hasDbXref ?doid ;
            oboInOwl:hasDbXref ?mondo_gard_id ;
            rdfs:label ?mondo_label .
        FILTER(CONTAINS(str(?doid), 'DOID'))
        FILTER(CONTAINS(str(?mondo_gard_id), 'GARD'))
    }"
) %>%
    tidy_sparql() %>%
    dplyr::mutate(
        gard_id = stringr::str_replace_all(
            mondo_gard_id,
            "GARD:[0]+([0-9]+)",
            "GARD:\\1"
        )
    ) %>%
    dplyr::relocate(gard_id, .before = mondo_gard_id)
```

```{r save_mondo, include = FALSE, eval = FALSE}
# save Mondo data, can only be run interactively
readr::write_tsv(mondo_bridge, here::here("data/mapping/mondo_bridge-DO_GARD.tsv"))
```


# GARD-DO Relationships

Identifying DO-GARD mappings is ordered by quality as follows:

1. Already in DO = GARD xrefs in DO
2. Likely in DO, but missing GARD xref
    A. DO-GARD xref matches in Mondo
    B. DO-GARD exact string matches
3. Maybe in DO = DO-GARD label approximate string matches


## GARD Already in DO (mappings/xrefs)

```{r already_in}
gard_in_do <- dplyr::filter(do, !is.na(gard_id)) %>%
    dplyr::left_join(dplyr::rename(gard, gard_label = label), by = "gard_id") %>%
    dplyr::mutate(gard_deprec = is.na(gard_label))
```

- DO diseases with GARD xref = `r format(dplyr::n_distinct(gard_in_do$doid), big.mark = ",")`
- GARD diseases in DO = `r format(dplyr::n_distinct(gard_in_do$gard_id), big.mark = ",")`
- Total DO-GARD mappings = `r format(nrow(gard_in_do), big.mark = ",")`


### Mappings in Need of Curation Review

A. Deprecated GARD xrefs that are in DO.

```{r do_gard_deprec}
do_gard_deprec <- gard_in_do %>%
    dplyr::filter(gard_deprec)

do_gard_deprec
```

```{r curate-do_gard_deprec, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(do_gard_deprec, gard_gs, "DO-GARD_deprec")
```


B. DO diseases with more than one GARD xref.

```{r do_multi_gard}
do_multi_gard <- gard_in_do %>%
    dplyr::filter(DO.utils::all_duplicated(doid)) %>%
    dplyr::arrange(doid, gard_id) %>%
    dplyr::mutate(gard_deprec = dplyr::na_if(gard_deprec, FALSE))

do_multi_gard
```

```{r curate-do_multi_gard, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(do_multi_gard, gard_gs, "DO-multi_GARD_xref")
```


C. GARD diseases linked to more than one DO disease.

```{r gard_dup}
gard_in_do_dup <- gard_in_do %>%
    dplyr::filter(DO.utils::all_duplicated(gard_id)) %>%
    dplyr::arrange(gard_id, doid) %>%
    dplyr::mutate(gard_deprec = dplyr::na_if(gard_deprec, FALSE))

gard_in_do_dup
```

```{r curate-gard_dup, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(gard_in_do_dup, gard_gs, "GARD_xref-multi_DO")
```

All three of these sets of DO-GARD xrefs **should be reviewed** and likely updated.


```{r do_gard_ok, include = FALSE}
do_gard_ok <- gard_in_do %>%
    dplyr::filter(
        !gard_id %in% do_gard_deprec$gard_id,
        !doid %in% do_multi_gard$doid,
        !gard_id %in% gard_in_do_dup$gard_id
    ) %>%
    dplyr::mutate(gard_deprec = dplyr::na_if(gard_deprec, FALSE))
```


```{r save-do_gard_ok, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(do_gard_ok, gard_gs, "DO-GARD_ok")
```

Total DO-GARD mappings not needing curation review: `r format(nrow(do_gard_ok), big.mark = ",")`.


## GARD Easily Added to DO

### GARD-DO Bridged Mappings (Mondo)

Using Mondo as a bridge, the following DO-GARD mappings may be correct:

```{r mondo_bridge}
mondo_do_gard <- mondo_bridge %>%
    dplyr::left_join(dplyr::select(do, doid, do_label = label), by = "doid") %>%
    dplyr::left_join(dplyr::rename(gard, gard_label = label), by = "gard_id") %>%
    dplyr::select(doid, do_label, gard_id, gard_label) %>%
    # remove deprecated GARD xrefs
    dplyr::filter(!is.na(gard_label)) %>%
    # remove DO-GARD mappings in DO, handled in [1]
    dplyr::mutate(do_gard = paste0(doid, gard_id)) %>%
    dplyr::filter(
        !do_gard %in% 
            dplyr::mutate(gard_in_do, do_gard =  paste0(doid, gard_id))$do_gard
    ) %>%
    dplyr::select(-do_gard) %>%
    # identify where Mondo suggests alternate GARD xrefs
    dplyr::mutate(
        suggesting_alt_gard = dplyr::na_if(doid %in% gard_in_do$doid, FALSE)
    ) %>%
    dplyr::arrange(doid, gard_id)

mondo_do_gard
```

```{r curate-mondo_bridge, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(mondo_do_gard, gard_gs, "DO-GARD_xref_bridge")
```

Total GARD xrefs that are likely in DO: `r nrow(mondo_do_gard)`.


```{r cum_do_gard1, include = FALSE}
do_potential_gard <- union(gard_in_do$doid, mondo_do_gard$doid)
potential_gard <- union(gard_in_do$gard_id, mondo_do_gard$gard_id)
```


### Exact GARD-DO Term Matches

String matching may identify other DO-GARD mapping candidates. String matching approaches are consist of exact matches and approximate (or fuzzy) matches.

For both approaches, comparison is made after initial standardization of disease labels (or synonyms), which consists of replacing all punctuation and extra spaces with a single space and converting words to lowercase. Numbers are _not_ standardized (e.g. roman numerals are not converted to normal numbers).

Exact matches specifically are also approached step-wise prioritizing matches between DO and GARD labels over GARD labels and DO synonyms.

#### Exact GARD-DO Label Matches

```{r label_std}
do_std <- do %>%
    dplyr::filter(!doid %in% do_potential_gard) %>%
    dplyr::mutate(
        label_std = label %>%
            stringr::str_to_lower() %>%
            stringr::str_replace_all(
                "([:blank:]|[:punct:]|[:space:])+",
                " "
            )
    ) %>%
    dplyr::select(-gard_id)

gard_std <- gard %>%
    dplyr::filter(!gard_id %in% potential_gard) %>%
    dplyr::mutate(
        label_std = label %>%
            stringr::str_to_lower() %>%
            stringr::str_replace_all(
                "([:blank:]|[:punct:]|[:space:])+",
                " "
            )
    )
```

```{r label_exact_match}
do_gard_exact_label <- do_std %>%
    dplyr::filter(label_std %in% gard_std$label_std) %>%
    dplyr::left_join(
        dplyr::rename(gard_std, gard_label = label),
        by = "label_std"
    ) %>%
    dplyr::mutate(match_type = "label-exact") %>%
    dplyr::relocate(doid:label, match_type, gard_id:gard_label, label_std)
```

DO and GARD have `r nrow(do_gard_exact_label)` exact label matches.

```{r show-label_exact, echo = FALSE}
do_gard_exact_label
```

```{r cum_do_gard2, include = FALSE}
do_potential_gard <- union(do_potential_gard, do_gard_exact_label$doid)
potential_gard <- union(potential_gard, do_gard_exact_label$gard_id)

do_std <- do_std %>%
    dplyr::filter(!doid %in% do_potential_gard)
gard_std <- gard_std %>%
    dplyr::filter(!gard_id %in% potential_gard)
```


#### Exact GARD Label-DO Synonym Matches

```{r std_syn}
do_syn_std <- do_syn %>%
    dplyr::mutate(
        synonym_std = synonym %>%
            stringr::str_to_lower() %>%
            stringr::str_replace_all(
                "([:blank:]|[:punct:]|[:space:])+",
                " "
            )
    ) %>%
    dplyr::filter(!duplicated(synonym_std))
```

```{r syn_exact_match}
do_gard_exact_syn <- do_syn_std %>%
    dplyr::filter(synonym_std %in% gard_std$label_std) %>%
    dplyr::left_join(
        dplyr::rename(gard_std, gard_label = label),
        by = c("synonym_std" = "label_std")
    ) %>%
    dplyr::mutate() %>%
    dplyr::group_by(doid) %>%
    dplyr::mutate(
        match_type = dplyr::if_else(
            length(doid) > 1,
            "synonym-exact-mult_match",
            "synonym-exact"
        )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::relocate(
        gard_id:gard_label, match_type, doid:synonym, term_std = synonym_std
    ) %>%
    dplyr::arrange(gard_id)
```

`r nrow(do_gard_exact_syn)` GARD terms have exact DO synonym matches. Some of these GARD terms match synonyms from more than one DO term.

```{r show-syn_exact, echo = FALSE}
do_gard_exact_syn
```

```{r cum_do_gard3, include = FALSE}
do_potential_gard <- union(do_potential_gard, do_gard_exact_syn$doid)
potential_gard <- union(potential_gard, do_gard_exact_syn$gard_id)

do_std <- do_std %>%
    dplyr::filter(!doid %in% do_potential_gard)
gard_std <- gard_std %>%
    dplyr::filter(!gard_id %in% potential_gard)
do_syn_std <- do_syn_std %>%
    dplyr::filter(!doid %in% do_potential_gard)
```


#### Summary: Exact Term Matches

```{r}
do_gard_exact <- dplyr::bind_rows(
    dplyr::rename(do_gard_exact_label, term_std = label_std),
    do_gard_exact_syn
) %>%
    dplyr::mutate(
        do_term = dplyr::if_else(is.na(label), synonym, label),
        do_term_type = dplyr::if_else(is.na(label), "synonym", "label")
    ) %>%
    dplyr::select(
        doid, do_term, do_term_type, match_type, gard_id:gard_label, term_std
    )
```

Total GARD-DO exact matches: `r format(dplyr::n_distinct(do_gard_exact$gard_id), big.mark = ",")`.

```{r curate-exact_match, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(do_gard_exact, gard_gs, "DO-GARD_exact_match")
```


## GARD Possibly Added to DO

Approximate string matches _can_ be very inaccurate, so thee results below  will need particularly _careful_ scrutiny.

For approximate string matching the 'longest common substring distance' matching algorithm is used with a maximum difference equal to the lower of the two string sets 3rd quartile of string length (rounded down).


### Approximate GARD-DO Label Matches

```{r approx_fn, include = FALSE}
max_dist <- function(x, y) {
    x_3q <- summary(stringr::str_length(x))[5]
    y_3q <- summary(stringr::str_length(y))[5]
    
    floor(min(c(x_3q, y_3q)))
}

add_match_score <- function(match_tbl, col_nm = "score") {
    dplyr::mutate(
        match_tbl,
        {{ col_nm }} := 1 - dist / 
            pmin(
                stringr::str_length(x),
                stringr::str_length(table_match)
            )
    )
}
```

```{r fz_label_prep}
label_dist <- max_dist(do_std$label_std, gard_std$label_std)
```

The maximum difference for comparison of the DO & GARD labels is `r label_dist`.

```{r fz_label_match}
system.time(
    gard_do_fz <- DO.utils::match_fz(
        gard_std$label_std,
        do_std$label_std,
        maxDist = label_dist
    )
)

gard_do_fz <- gard_do_fz %>%
    add_match_score() %>%
    dplyr::rename(gard_std = x, do_std = table_match) %>%
    dplyr::arrange(dplyr::desc(score))
```

```{r show-fz_label, echo = FALSE}
gard_do_fz
```


### Approximate GARD Label-DO Synonym Matches

```{r fz_syn_prep}
syn_dist <- max_dist(do_syn_std$synonym_std, gard_std$label_std)
```

The maximum difference for comparison of the DO & GARD labels is `r syn_dist`.

```{r fz_syn_match}
system.time(
    gard_do_fz_syn <- DO.utils::match_fz(
        gard_std$label_std,
        do_syn_std$synonym_std,
        maxDist = syn_dist
    )
)

gard_do_fz_syn <- gard_do_fz_syn %>%
    add_match_score() %>%
    dplyr::rename(gard_std = x, do_syn_std = table_match) %>%
    dplyr::arrange(dplyr::desc(score))
```

```{r show-fz_syn, echo = FALSE}
gard_do_fz_syn
```

### Summary: Approximate Term Matches

```{r fz_all}
# prep for merge - add DO info for each dataset
gard_do_fz <- gard_do_fz %>%
    dplyr::left_join(
        dplyr::select(do_std, doid, do_std = label_std),
        by = "do_std"
    ) %>%
    dplyr::mutate(do_term_type = "label")

gard_do_fz_syn <- gard_do_fz_syn %>%
    dplyr::rename(do_std = do_syn_std) %>%
    dplyr::left_join(
        dplyr::select(do_syn_std, doid, do_std = synonym_std),
        by = "do_std"
    ) %>%
    dplyr::mutate(do_term_type = "synonym")

# merge
fz_all <- dplyr::bind_rows(
    gard_do_fz,
    gard_do_fz_syn
) %>%
    # add GARD info
    dplyr::left_join(
        dplyr::select(gard_std, gard_id, gard_std = label_std),
        by = "gard_std"
    ) %>%
    # drop term type from non-matches & drop duplicates
    dplyr::mutate(
        do_term_type = dplyr::if_else(
            is.na(doid),
            NA_character_,
            do_term_type
        )
    ) %>%
    unique() %>%
    # sort by GARD matching group with best scoring groups first and within
    #   group
    dplyr::group_by(gard_id) %>%
    dplyr::mutate(
        best_score = ifelse(
            all(is.na(score)),
            NA,
            max(score, na.rm = TRUE)
        )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(
        dplyr::desc(best_score),
        gard_id,
        dplyr::desc(score),
        do_term_type
    ) %>%
    dplyr::mutate(
        gard_group = cumsum(
            gard_id != tidyr::replace_na(dplyr::lag(gard_id), "")
        )
    ) %>%
    dplyr::relocate(
        gard_group, gard_id, gard_std, score, dist, doid, do_std, do_term_type
    ) %>%
    dplyr::rename(chr_diff = dist)
```

Drop matches below a score of 0.6. Quick manual evaluation suggests this cutoff retains most real and related hits, while dropping the majority of totally unrelated results. A sizable number of poor hits are retained with the goal of being more inclusive.

```{r fz_quality}
# discard poor matches (being generous here in thos
fz_gt60 <- dplyr::filter(fz_all, score >= 0.6)
```

This dataset includes approximate matches from DO for `r dplyr::n_distinct(fz_gt60$gard_id)` GARD diseases.

```{r curate-fz_match, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(fz_gt60, gard_gs, "GARD-DO_fz_match")
```


# Summary: Overall

```{r GARD_no_DO, include = FALSE}
gard_no_do <- gard %>%
    dplyr::filter(!gard_id %in% c(gard_in_do$gard_id, do_gard_exact$gard_id, fz_gt60$gard_id))
```

```{r save-GARD_no_DO, include = FALSE, eval = FALSE}
googlesheets4::write_sheet(gard_no_do, gard_gs, "GARD_no_match")
```


Disease Set         | count
:--|:--
All GARD diseases   | `r format(dplyr::n_distinct(gard$gard_id), big.mark = ",")`
   |  
_GARD diseases in DO (need review)_* | `r dplyr::n_distinct(gard_in_do$gard_id) - dplyr::n_distinct(do_gard_ok$gard_id)`
_GARD diseases in DO (okay)_ | `r format(dplyr::n_distinct(do_gard_ok$gard_id), big.mark = ",")`
**TOTAL** GARD diseases in DO | `r format(dplyr::n_distinct(gard_in_do$gard_id), big.mark = ",")`
   |  
GARD diseases with strong matches in DO | `r dplyr::n_distinct(do_gard_exact$gard_id)`
**TOTAL** in DO + strong matches | `r format(dplyr::n_distinct(c(gard_in_do$gard_id, do_gard_exact$gard_id)), big.mark = ",")`
   |  
GARD diseases with approximate matches in DO | `r dplyr::n_distinct(fz_gt60$gard_id)`
**TOTAL** in DO + any matches | `r format(dplyr::n_distinct(c(gard_in_do$gard_id, do_gard_exact$gard_id, fz_gt60$gard_id)), big.mark = ",")`
   |  
Remaining GARD diseases | `r format(dplyr::n_distinct(gard_no_do$gard_id), big.mark = ",")`

*Some GARD diseases have been deprecated, resulting in slight adding discrepancies.

