---
title: "Produce Candidate Definitions for DOIDs without them"
author: "J. Allen Baron"
date: "2021-11-02"
output:
    html_document:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

```{r setup, include = FALSE}
library(here)
library(virtuoso)
library(parallel)
library(tidyverse)
library(keyring)
library(rentrez)
library(googlesheets4)

options(gargle_oauth_email = "allenbaron@som.umaryland.edu")
```

# PURPOSE
To obtain candidate definitions from MESH and NCI Thesaurus for DOIDs without definitions.


# Data Setup (Manual Execution)

## Download Data

1. Copy doid.owl (release: v2021-10-11) from HumanDiseaseOntology git repo to data.
2. Download NCI Thesaurus OWL file (asserted version; release 21.09d) to data from https://evs.nci.nih.gov/evs-download/thesaurus-downloads.
    - NOTE: .zip can't be read by `virtuoso` bulk database uploader, so I unzipped and recompressed it as .gz file.
3. Download MESH N-triples file to data from https://hhs.github.io/meshrdf/ ftp site (release: 2021-10-28, updated nightly).

```{r}
# data files
doid_owl <- here::here("data/doid_v2021-10-11.owl")
nci_owl <- here::here("data/Thesaurus.owl.gz")
mesh_nt <- here::here("data/mesh.nt.gz")
```

```{r eval = FALSE, include = FALSE}
# extra if using NCI thesaurus flat file
nci_tsv <- here::here("data/Thesaurus_21.09d.FLAT.zip")
nci_colnames <- c("code", "concept_name", "parents", "synonyms", "definition",
                  "display_name", "concept_status", "semantic_type")
nci <- readr::read_tsv(nci_tsv, col_names = nci_colnames)
```


## Setup a Virtuoso Database

I installed Virtuoso using `scripts/setup_virtuoso.R` and default values. That script and SPARQL queries herein rely on the `virtuoso` R package.

I chose to load the 3 rdf files as individual graphs so that I could explore them separately and avoid unexpected results from queries finding hits in multiple data sources.

```{r message = FALSE}
# start database and initiate a connection
virtuoso::vos_start()
vcon <- virtuoso::vos_connect()
```

```{r message = FALSE}
# list graphs (i.e. URIs)
vgraphs <- virtuoso::vos_list_graphs(vcon)$g

# import data into db, ran interactively once
if (!"doid" %in% vgraphs) {
    virtuoso::vos_import(
        con = vcon,
        files = doid_owl,
        graph = "doid",
        n_cores = dplyr::if_else(
            parallel::detectCores() > 2,
            parallel::detectCores() - 2,
            1
        )
    )
}

if (!"mesh" %in% vgraphs) {
    virtuoso::vos_import(
        con = vcon,
        files = mesh_nt,
        graph = "mesh",
        n_cores = dplyr::if_else(
            parallel::detectCores() > 2,
            parallel::detectCores() - 2,
            1
        )
    )
}

if (!"nci" %in% vgraphs) {
    virtuoso::vos_import(
        con = vcon,
        files = nci_owl,
        graph = "nci",
        n_cores = dplyr::if_else(
            parallel::detectCores() > 2,
            parallel::detectCores() - 2,
            1
        )
    )
}
```


# DOIDs without definitions

I extracted the DOIDs without definitions, excluding obsolete terms, along with all annotated xrefs for further analysis. I did not use the `src/sparql/extra/DO_no_defs.rq` SPARQL query (it has the same terms but results in multiple rows per term because it represents parents). Instead, I directly excluded obsolete terms and did not extract parent information.
```{r}
query_no_def_xref <-
    '# Query non-deprecated DOIDs without definitions
    PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
    PREFIX owl: <http://www.w3.org/2002/07/owl#>
    PREFIX obo: <http://purl.obolibrary.org/obo/>
    PREFIX oboInOwl: <http://www.geneontology.org/formats/oboInOwl#>

    SELECT ?doid ?do_name ?xref
    FROM <doid>
    WHERE {
        ?class a owl:Class ;
            oboInOwl:id ?doid ;
            rdfs:label ?do_name .
        OPTIONAL { ?class oboInOwl:hasDbXref ?xref . }
        # include only those missing definitions & not deprecated
        FILTER NOT EXISTS { ?class obo:IAO_0000115 ?def . }
        FILTER NOT EXISTS { ?class owl:deprecated ?any . }
    }'

do_no_def <- virtuoso::vos_query(vcon, query_no_def_xref) %>%
    tibble::as_tibble() %>%
    dplyr::mutate(xref_ns = stringr::str_remove(xref, ":.*"))
```


## General Statistics

There are **`r format(dplyr::n_distinct(do_no_def$doid), big.mark = ",")` DOIDs without definitions**. Of these, `r dplyr::n_distinct(do_no_def$doid[is.na(do_no_def$xref)])` have NO xrefs.

Here's a quick count of the xrefs generally:
```{r}
dplyr::count(do_no_def, xref_ns) %>%
    dplyr::arrange(dplyr::desc(n))
```

```{r}
nci_mesh_umls <- do_no_def %>%
    dplyr::filter(xref_ns %in% c("MESH", "NCI", "UMLS_CUI")) %>%
    dplyr::group_by(doid) %>%
    dplyr::summarize(
        nci_mesh = any(xref_ns %in% c("MESH", "NCI")),
        umls = any(xref_ns == "UMLS_CUI")
    )

nm_no_umls <- nci_mesh_umls %>%
    dplyr::filter(nci_mesh & !umls)
```

And lastly, the number of DOIDs without definitions **with or without NCI/MESH xrefs** is:
```{r}
nci_mesh_xref <- do_no_def %>%
    dplyr::group_by(doid) %>%
    dplyr::summarize(
        has_nci_mesh_xref = any(xref_ns %in% c("MESH", "NCI"))
    )

nci_mesh_xref %>%
    dplyr::count(has_nci_mesh_xref)
```

That's about 1/5 of the DOIDs without definitions that NCI & MESH can't help with. Among those it appears most have xrefs to one of the ICD's or SNOMED (via UMLS). I'm not sure any of those will be accessible for definitions.
```{r}
doid_no_nci_mesh <- dplyr::filter(nci_mesh_xref, !has_nci_mesh_xref)$doid

do_no_nci_mesh <- do_no_def %>%
    dplyr::filter(doid %in% doid_no_nci_mesh)

do_no_nci_mesh %>%
    dplyr::count(xref_ns) %>%
    dplyr::arrange(dplyr::desc(n))
```


## [Potentially] Actionable DOID w/o Def Lists

### Need xrefs

Here are the `r dplyr::n_distinct(do_no_def$doid[is.na(do_no_def$xref)])` DOIDs without definitions that have NO xrefs.:
```{r}
dplyr::filter(do_no_def, is.na(xref)) %>%
    dplyr::select(-xref_ns)
```


### [Potentially] Missing UMLS xrefs

I assumed that most NCI and MESH terms are also in UMLS. Of the `r format(dplyr::n_distinct(nci_mesh_umls$doid[nci_mesh_umls$nci_mesh]), big.mark = ",")` DOIDs with NCI or MESH xrefs there are `r format(dplyr::n_distinct(nm_no_umls$doid), big.mark = ",")` DOIDs _WITHOUT_ UMLS xrefs. Those DOIDs are:
```{r}
dplyr::filter(do_no_def, doid %in% nm_no_umls$doid) %>%
    dplyr::group_by(doid, do_name) %>%
    dplyr::summarize(xref = paste0(xref, collapse = "|"), .groups = "drop")
```


### [Potentially] Get Defs from EFO/GARD/OMIM

The small number of DOIDs without definitions with xrefs to EFO/GARD/OMIM are:
```{r}
dplyr::filter(do_no_nci_mesh, xref_ns %in% c("EFO", "GARD", "OMIM")) %>%
    dplyr::group_by(doid, do_name) %>%
    dplyr::summarize(xref = paste0(xref, collapse = "; "), .groups = "drop")
```


# Extraction of Definitions from NCI & MESH

## SPARQL Queries to RDF Files

I attempted to extract definitions from the MESH & NCI rdf files. Though the SPARQL query works, only NCI definitions are returned. It appears that there are _NO_ `meshv:scopeNote` (definition) triples for the MESH terms of interest in the file.
```{r}
query_new_def <-
    '# Query non-deprecated DOIDs without definitions
    PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
    PREFIX owl: <http://www.w3.org/2002/07/owl#>
    PREFIX obo: <http://purl.obolibrary.org/obo/>
    PREFIX oboInOwl: <http://www.geneontology.org/formats/oboInOwl#>
    PREFIX nci: <http://ncicb.nci.nih.gov/xml/owl/EVS/Thesaurus.owl#>
    PREFIX mesh: <http://id.nlm.nih.gov/mesh/>
    PREFIX meshv: <http://id.nlm.nih.gov/mesh/vocab#>

    SELECT ?doid ?do_name ?def ?xref
    FROM <doid>
    FROM <mesh>
    FROM <nci>
    WHERE {
        {
            SELECT *
            FROM <doid>
            FROM <mesh>
            WHERE {
                ?class a owl:Class ;
                    oboInOwl:id ?doid ;
                    rdfs:label ?do_name ;
                    oboInOwl:hasDbXref ?xref .
                FILTER NOT EXISTS { ?class obo:IAO_0000115 ?any_def . }
                FILTER NOT EXISTS { ?class owl:deprecated ?any . }

                # MESH defs
                FILTER STRSTARTS(STR(?xref), "MESH")
                BIND(
                    IRI(CONCAT(
                        "http://id.nlm.nih.gov/mesh/",
                        STRAFTER( STR(?xref), ":" )
                    )) AS ?uri
                )
                OPTIONAL {
                    # ?uri meshv:scopeNote ?def .       # no results
                    # ?uri ?p ?o .                      # has results but nothing
                    # ?uri meshv:identifier ?code .     #   resmbling definitions
                }
            }
        }
        UNION
        {
            SELECT *
            FROM <doid>
            FROM <nci>
            WHERE {
                ?class a owl:Class ;
                    oboInOwl:id ?doid ;
                    rdfs:label ?do_name ;
                    oboInOwl:hasDbXref ?xref .
                FILTER NOT EXISTS { ?class obo:IAO_0000115 ?any_def . }
                FILTER NOT EXISTS { ?class owl:deprecated ?any . }

                # NCI defs
                FILTER STRSTARTS(STR(?xref), "NCI")
                BIND( STRAFTER( STR(?xref), ":" ) AS ?code )
                OPTIONAL {
                    ?uri nci:P97 ?def ;
                        nci:NHC0 ?code .
                    # remove obsolete/retired concepts from NCI
                    FILTER NOT EXISTS {
                        ?uri nci:P310 ?old_concept .
                        VALUES ?old_concept { "Obsolete_Concept" "Retired_Concept" }
                    }
                }
            }
        }
    }'

do_nci <- virtuoso::vos_query(vcon, query_new_def) %>%
    tibble::as_tibble()
```

I attempted to identify a definition (`meshv:scopeNote`) for a specific MESH record that I know has one (MESH:D010211, "Papilledema", https://meshb.nlm.nih.gov/record/ui?ui=D010211), by submitting queries to MeSH's SPARQL API endpoint. As in my attempts to get definitions from the mesh.nt.gz file download, there is no `meshv:scopeNote` triple available via the SPARQL API endpoint, with or without inference on.

<details>
<summary>SPARQL query submitted to the API:</summary>

```
PREFIX rdf: <http://www.w3.org/1999/02/22-rdf-syntax-ns#>
PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>
PREFIX xsd: <http://www.w3.org/2001/XMLSchema#>
PREFIX owl: <http://www.w3.org/2002/07/owl#>
PREFIX meshv: <http://id.nlm.nih.gov/mesh/vocab#>
PREFIX mesh: <http://id.nlm.nih.gov/mesh/>
PREFIX mesh2021: <http://id.nlm.nih.gov/mesh/2021/>
PREFIX mesh2020: <http://id.nlm.nih.gov/mesh/2020/>
PREFIX mesh2019: <http://id.nlm.nih.gov/mesh/2019/>

SELECT *
FROM <http://id.nlm.nih.gov/mesh>
WHERE {
  ?class ?p ?o .
  VALUES ?class { mesh:D010211 }
}
```

<details>

I have not been able to identify a downloadable or queryable source, besides the MESH rdf resources, from which to get definitions. The last option I can think of is to try the Entrez API.


## MESH Entrez API Queries

I extracted the MESH codes and checked to find out what access to MESH the Entrez API has.
```{r}
mesh_codes <- do_nci %>%
    dplyr::filter(stringr::str_detect(xref, "MESH")) %>%
    .$xref %>%
    stringr::str_remove("MESH:")
```

MESH is an available database and is searchable by code (i.e. NLM MeSH Browser Unique ID).
```{r include = FALSE}
# get ENTREZ API key via keyring
rentrez::set_entrez_key(keyring::key_get("ENTREZ_KEY"))
```

```{r}
rentrez::entrez_dbs()
rentrez::entrez_db_summary("mesh")
rentrez::entrez_db_searchable("mesh")
```

Hopefully, the definitions available in the online MeSH search are also accessible via the Entrez API.

I thought the definitions would most likely be included in the summary info so I attempted to access the summaries but nothing was returned. After some investigation I discovered the Entrez API is using a different internal unique identifier for their records (UID instead of MHUI; not visible on search result pages).

Online searches didn't work, so I used an Entrez API search with MHUIs to get the UIDs, and then was able to obtain the MeSH summary information, which included Scope Notes (ie. definitions!!).
```{r}
mc_n <- length(mesh_codes)
mesh_wh_list <- purrr::map(
    seq(0, 800, by = 100),
    function(s) {
        s1 <- s + 1
        e <- dplyr::if_else(s + 100 < mc_n, s + 100, as.double(mc_n))

        rentrez::entrez_search(
            db = "mesh",
            term = paste0(paste0(mesh_codes[s:e], "[MHUI]"), collapse = " OR "),
            use_history = TRUE,
            retmax = 1000,
            retmode = "xml"
        )
    }
)
```

```{r}
mesh_uid <- purrr::map(mesh_wh_list, ~.x$ids) %>%
    unlist()
mesh_wh <- rentrez::entrez_post(db = "mesh", id = mesh_uid)

# avoid extra, unnecessary API calls when knitting
mesh_summary_file <- here::here("data/mesh_summary-20211102.rda")
if (!file.exists(mesh_summary_file)) {
    mesh_summary <- rentrez::entrez_summary(
        db = "mesh",
        web_history = mesh_wh,
        always_return_list = TRUE,
        retmode = "xml"
    )
    save(mesh_summary, file = mesh_summary_file)
} else {
    load(mesh_summary_file)
}

mesh_tbl <- mesh_summary %>%
        tibble::enframe(name = "esummary_id", value = "tmp") %>%
        tidyr::unnest_wider(col = "tmp")

mesh_tbl_tidy <- mesh_tbl %>%
    dplyr::mutate(xref = paste0("MESH:", DS_MeSHUI)) %>%
    dplyr::select(def = DS_ScopeNote, xref)
```

Some of the NLM MeSH Browser Unique IDs (MHUI) I passed to the API did not return Unique IDs (UIDs; `r nrow(mesh_tbl)` records for `r length(mesh_codes)` MHUIs were returned). I checked to see which MHUIs did not have matches and found none.
```{r}
mesh_codes[!mesh_codes %in% mesh_tbl$DS_MeSHUI]
```

After a little more exploration, I discovered that I'd included duplicates in the list to the API and that there are _not_ actually any missing. So the data looks complete! ... in the sense that there is a record for every xref I have.

I merged this MESH data in with the DOIDs and NCI definitions and cleaned it up a bit to make it more human-readable.
```{r}
do_nci_mesh <- dplyr::left_join(do_nci, mesh_tbl_tidy, by = "xref") %>%
    # combine definition cols
    dplyr::mutate(def = dplyr::if_else(!is.na(def.x), def.x, def.y)) %>%
    dplyr::select(-def.x, -def.y) %>%
    unique() %>%
    dplyr::arrange(doid, dplyr::desc(xref))
```


## Xref Definition Statistics

For those DOIDs with NCI and/or MeSH xrefs, the overall count of DOIDs with candidate definitions from both NCI & MeSH, just one of them, or none is:
```{r include = FALSE}
# custom functions
mesh_present <- function(x, d) {
    any(stringr::str_detect(x, "MESH") & !is.na(d))
}
nci_present <- function(x, d) {
    any(stringr::str_detect(x, "NCI") & !is.na(d))
}
```

```{r}
do_nci_mesh %>%
    dplyr::group_by(doid) %>%
    dplyr::summarize(
        def_source = dplyr::case_when(
             mesh_present(xref, def) & nci_present(xref, def) ~ "NCI & MESH",
             nci_present(xref, def) ~ "NCI",
             mesh_present(xref, def) ~ "MESH",
             all(is.na(def)) ~ "none"
        )
    ) %>%
    dplyr::mutate(
        def_source = factor(def_source, levels = c("NCI & MESH", "NCI", "MESH",
                                                 "none"))
    ) %>%
    dplyr::count(def_source, name = "doid_n")
```

The count of NCI/MeSH xrefs without their own definitions is:
```{r}
xref_no_def <- do_nci_mesh %>%
    dplyr::filter(is.na(def) & !is.na(xref)) %>%
    unique()

xref_no_def %>%
    dplyr::mutate(xref_ns = stringr::str_remove(xref, ":.*")) %>%
    dplyr::count(xref_ns)
```

And those xrefs without any definitions are:
```{r}
xref_no_def %>%
    dplyr::arrange(doid, dplyr::desc(xref))
```


# Save Candidate Definitions

Lastly I saved the definitions data to a Google Sheets file in `Disease_Ontology/DO_definitions/Sets of terms to Define/2021-Missing_Definitions`.
```{r}
gs_id <- googlesheets4::as_sheets_id(
    "https://docs.google.com/spreadsheets/d/1aHfmvsO5BBPTIWL_mUcxZyEAEJDuoDVrWmbAITzH2MQ/edit?usp=sharing"
)
googlesheets4::write_sheet(data = do_nci_mesh, ss = gs_id, sheet = "definitions")
```
