---
title: "Analysis of RTT abbreviation"
author: "J. Allen Baron"
date: "2025-09-19"
format: 
    html:
        toc: true
        toc_float: true
        code_folding: hide
    gfm: default
    pdf: default
---

Seems to be used frequently for Rett syndrome (DOID:1206), but should it be added?

**GOAL:** Determine if RTT is worth adding as a synonym of Rett syndrome.

_It will be if it is not used for other diseases and Rett syndrome appears to be a sizable proportion of instances identified in the published literature._


```{r setup, include=FALSE}
library(europepmc)
library(tidyverse)
library(xml2)
library(here)

knitr::opts_chunk$set(echo = if (knitr::is_html_output()) TRUE else FALSE)
```

```{r in_progress_data}
data_file <- here::here("data/disease_info/RTT-abbrev.rda")
```

```{r custom_functions}
# get_ftxt_safely() will automatically get PMC articles or books and will NOT
#   fail if on errors caused by individual download failures
safe_epmc_ftxt <- purrr::safely(europepmc::epmc_ftxt, otherwise = NA, quiet = FALSE)
safe_epmc_ftxt_bk <- purrr::safely(europepmc::epmc_ftxt_book, otherwise = NA, quiet = FALSE)

get_ftxt_safely <- function(pmcid = NA, bookid = NA) {
    if (is.na(pmcid) && is.na(bookid) ) return(NA)
    if (!is.na(pmcid)) {
        out <- list(safe_epmc_ftxt(pmcid))
    } else {
        out <- list(safe_epmc_ftxt_bk(bookid))
    }
    cat(".")
    out
}

# drops retractions & pre-prints, reduces columns, and formats dates
tidy_epmc_search <- function(.df) {
    tidy_df <- .df |>
        dplyr::filter(!stringr::str_detect(pubType, "retract|preprint")) |>
        dplyr::select(
            "id", "title", "pubYear", pubDate = "firstPublicationDate",
            "pmcid", dplyr::any_of("bookid")
        ) |> 
        dplyr::mutate(
            pubDate = lubridate::as_date(pubDate),
            pubYear = lubridate::year(pubDate)
        )
    # bookid sometimes missing --> standardize
    if (!("bookid" %in% names(tidy_df))) {
        tidy_df <- tidy_df |>
            dplyr::mutate(bookid = NA_character_)
    }
    tidy_df
}

get_open_access <- function(tidy_epmc, nmax = 5000) {
    # limit to open access likely to be available
    oa <- tidy_epmc |>
        dplyr::filter(isOpenAccess == "Y" & !(is.na(pmcid) & is.na(bookid)))
    
    # sample maintaining general pattern over time, if too many
    oa_total <- nrow(oa)
    if (oa_total > nmax) {
        warning(
            paste0(
                oa_total, " open access publications found; ",
                "sampling ≤ ", nmax
            ),
            immediate. = TRUE
        )
        n_yr <- dplyr::count(oa, pubYear, name = "yr_total") |>
            dplyr::mutate(
                frac = .data$yr_total / sum(.data$yr_total),
                frac_n = DO.utils::round_down(.data$frac * 5000)
            )
        oa <- purrr::map2(
            n_yr$pubYear,
            n_yr$frac_n,
            ~ oa |> 
                dplyr::filter(.data$pubYear == .x) |>
                dplyr::slice_sample(n = .y)
        ) |>
            dplyr::bind_rows()
    }
    
    full_text <- oa |>
        dplyr::rowwise() |>
        dplyr::mutate(ft_xml = get_ftxt_safely(.data$pmcid, .data$bookid)) |>
        dplyr::mutate(ft = parse_ftxt_xml(ft_xml, "//body"))

    full_text
}

# parse_ftxt_xml() parses results from get_ftxt_safely()
parse_ftxt_xml <- function(safe_ftxt_xml, xml_accessor) {
    if (!is.null(safe_ftxt_xml$error)) {
        return(paste0("ERROR: ", safe_ftxt_xml$error$message))
    }
    out <- safe_ftxt_xml$result |>
        xml2::xml_find_all(xml_accessor) |>
        xml2::xml_text()

    if (length(out) == 0) {
        out <- paste0(
            "ERROR [NO BODY]: ",
            xml2::xml_text(safe_ftxt_xml$result)
        )
        if (length(out) == 0) {
            out <- "ERROR: No text extractable"
        }
    } else if (length(out) > 1) {
        out <- DO.utils::vctr_to_string(out, delim = "%%%%%") |>
            paste0("WARNING: Multilength output, separated by %%%%%.")
    }

    out
}
```


# Approach

In this case, I'll identify all open access publications with "RTT" in them and then try to extract the full-length equivalent from each publication to estimate the prevalence and determine how this abbreviation is used.

_The output will be saved to a file to avoid need of repeat API call (until the next desired updated)._

```{r}
abbrev <- "RTT"
search_str <- DO.utils::sandwich_text(abbrev, c('OPEN_ACCESS:y AND "', '"'))

if (!file.exists(data_file)) {
    res <- europepmc::epmc_search(search_str, synonym = FALSE, limit = 20000)

    save(res, file = data_file)
} else {
    load(data_file)
}
```

The number of publication hits was `r format(nrow(res), big.mark = ",")`.

The full text of Open Access publications will also be downloaded for this analysis, excluding retractions and preprints.

```{r}
use_bookid <- "bookid" %in% names(res)

res_tidy <- res |>
    dplyr::filter(
        !stringr::str_detect(.data$pubType, "retract|preprint"),
        !(is.na(.data$pmcid) & is.na(.data$bookid))
    ) |>
    dplyr::select(
        "id", "title", "pubYear", pubDate = "firstPublicationDate",
        "pmcid", if (use_bookid) { "bookid" }) |>
    dplyr::mutate(
        pubDate = lubridate::as_date(.data$pubDate),
        pubYear = lubridate::year(.data$pubDate)
    )

if (!exists("res_ftxt")) {
    res_ftxt <- res_tidy |>
        dplyr::rowwise() |>
        dplyr::mutate(
            ft_xml = get_ftxt_safely(
                .data$pmcid,
                if (use_bookid) { .data$bookid } else { NA }
            )
        ) |>
        dplyr::mutate(ft = parse_ftxt_xml(.data$ft_xml, "//body"))
    
    save(res, res_ftxt, file = data_file)
}
```


# Evaluating Usage

Extracting all these values from the full text of the sample publications and all the titles (in a case-insensitive manner).
```{r}
regex_str <- "\\b[Rr]([^ ]+ ){1,3}\\(RTT\\)"

anal_df <- res_ftxt |>
    dplyr::select(-"ft_xml") |>
    dplyr::mutate(
        ft_status = dplyr::case_when(
            .data$ft == "" ~ "none",
            stringr::str_detect(.data$ft, "^ERROR.*NO BODY") ~ "abstract?",
            stringr::str_detect(.data$ft, "^ERROR") ~ "inaccessible",
            !is.na(.data$ft) ~ "available"
        ),
        title_match = stringr::str_extract_all(
            .data$title,
            stringr::regex(regex_str, ignore_case = TRUE)
        ),
        ft_match = stringr::str_extract_all(
            .data$ft,
            stringr::regex(regex_str, ignore_case = TRUE)
        )
    ) |>
    tidyr::unnest(title_match, keep_empty = TRUE) |>
    tidyr::unnest(ft_match, keep_empty = TRUE) |>
    tidyr::pivot_longer(
        cols = c("title_match", "ft_match"),
        names_sep = "_",
        names_to = c("type", ".value")
    ) |>
    dplyr::mutate(
        type = dplyr::if_else(
            .data$type == "ft",
            paste(.data$type, .data$ft_status, sep = "-"),
            .data$type
        )
    ) |>
    dplyr::select(-"ft", -"ft_status")
```

The number of publications with and without matches in their titles or full text (or what approximates for full text) are as follows:

```{r}
anal_df |>
    dplyr::summarize(match = any(!is.na(.data$match)), .by = c("id", "type")) |>
    dplyr::count(.data$type, .data$match) |>
    dplyr::mutate(title = .data$type == "title") |>
    dplyr::mutate(pct = round(.data$n / sum(.data$n) * 100, 2), .by = "title") |>
    dplyr::arrange(.data$title, .data$match, .data$type) |>
    dplyr::select(-"title")
```
The breakdown of terms identified is as follows:

```{r}
terms <- anal_df |>
    dplyr::filter(!is.na(.data$match)) |>
    dplyr::mutate(
        std_match = stringr::str_remove(.data$match, "\\(.*") |>
            stringr::str_replace_all("[[:punct:]]", " ") |>
            stringr::str_to_lower()
    )

dplyr::count(terms, .data$std_match, sort = T) |>
    dplyr::mutate(pct = round(.data$n / sum(.data$n) * 100, 2))
```

# Conclusion

Though Rett syndrome represents about 1/3 of the uses for RTT, it is the major meaning for RTT in the disease realm. It seems reasonable to add RTT as a synonym for Rett syndrome (this will use the acronym annotation, since DO does not use an annotation for abbreviation, even though it's not exactly an acronym).
