---
title: "Analysis of CKM syndrome name & synonyms"
date: "2025-02-21"
output:
    html_notebook:
        toc: true
        toc_float: true
        code_folding: hide
---

This is a new disease requested by Sue Bello in HumanDiseaseOntology Issue [#1473](https://github.com/diseaseOntology/humanDiseaseOntology/issues/1473) and initially defined by the American Heart Association in 2023 (Ndumele, Chiadi E et al. “Cardiovascular-Kidney-Metabolic Health: A Presidential Advisory From the American Heart Association.” Circulation vol. 148,20 (2023): 1606-1635. doi:10.1161/CIR.0000000000001184; [PMID:37807924](https://pubmed.ncbi.nlm.nih.gov/37807924/).)

The goal is to run this analysis at various times to determine what synonyms to include for this disease. The AHA clearly stated that it should be named and referred to as a syndrome but authors/clinicians have a tendency to be free with the word disease for syndromic presentations, particularly for grouping terms (e.g. chronic kidney disease and cardiovascular disease). So the appearance of this disease and synonyms for it should be monitored.


```{r setup, include=FALSE}
library(europepmc)
library(tidyverse)
library(xml2)
library(here)
library(hues)
library(plotly)
```

```{r in_progress_data}
data_file <- here::here("data/disease_info/CKM_syndrome.rda")
```

```{r custom_functions}
# get_ftxt_safely() will automatically get PMC articles or books and will NOT
#   fail if on errors caused by individual download failures
safe_epmc_ftxt <- purrr::safely(europepmc::epmc_ftxt, otherwise = NA, quiet = FALSE)
safe_epmc_ftxt_bk <- purrr::safely(europepmc::epmc_ftxt_book, otherwise = NA, quiet = FALSE)

get_ftxt_safely <- function(pmcid = NA, bookid = NA) {
    if (is.na(pmcid) && is.na(bookid) ) return(NA)
    if (!is.na(pmcid)) {
        out <- list(safe_epmc_ftxt(pmcid))
    } else {
        out <- list(safe_epmc_ftxt_bk(bookid))
    }
    cat(".")
    out
}

# drops retractions & pre-prints, reduces columns, and formats dates
tidy_epmc_search <- function(.df) {
    tidy_df <- .df |>
        dplyr::filter(!stringr::str_detect(pubType, "retract|preprint")) |>
        dplyr::select(
            "id", "title", "pubYear", pubDate = "firstPublicationDate",
            "pmcid", dplyr::any_of("bookid")
        ) |> 
        dplyr::mutate(
            pubDate = lubridate::as_date(pubDate),
            pubYear = lubridate::year(pubDate)
        )
    # bookid sometimes missing --> standardize
    if (!("bookid" %in% names(tidy_df))) {
        tidy_df <- tidy_df |>
            dplyr::mutate(bookid = NA_character_)
    }
    tidy_df
}

get_open_access <- function(tidy_epmc, nmax = 5000) {
    # limit to open access likely to be available
    oa <- tidy_epmc |>
        dplyr::filter(isOpenAccess == "Y" & !(is.na(pmcid) & is.na(bookid)))
    
    # sample maintaining general pattern over time, if too many
    oa_total <- nrow(oa)
    if (oa_total > nmax) {
        warning(
            paste0(
                oa_total, " open access publications found; ",
                "sampling ≤ ", nmax
            ),
            immediate. = TRUE
        )
        n_yr <- dplyr::count(oa, pubYear, name = "yr_total") |>
            dplyr::mutate(
                frac = .data$yr_total / sum(.data$yr_total),
                frac_n = DO.utils::round_down(.data$frac * 5000)
            )
        oa <- purrr::map2(
            n_yr$pubYear,
            n_yr$frac_n,
            ~ oa |> 
                dplyr::filter(.data$pubYear == .x) |>
                dplyr::slice_sample(n = .y)
        ) |>
            dplyr::bind_rows()
    }
    
    full_text <- oa |>
        dplyr::rowwise() |>
        dplyr::mutate(ft_xml = get_ftxt_safely(.data$pmcid, .data$bookid)) |>
        dplyr::mutate(ft = parse_ftxt_xml(ft_xml, "//body"))

    full_text
}

# parse_ftxt_xml() parses results from get_ftxt_safely()
parse_ftxt_xml <- function(safe_ftxt_xml, xml_accessor) {
    if (!is.null(safe_ftxt_xml$error)) {
        return(paste0("ERROR: ", safe_ftxt_xml$error$message))
    }
    out <- safe_ftxt_xml$result |>
        xml2::xml_find_all(xml_accessor) |>
        xml2::xml_text()

    if (length(out) == 0) {
        out <- paste0(
            "ERROR [NO BODY]: ",
            xml2::xml_text(safe_ftxt_xml$result)
        )
        if (length(out) == 0) {
            out <- "ERROR: No text extractable"
        }
    } else if (length(out) > 1) {
        out <- DO.utils::vctr_to_string(out, delim = "%%%%%") |>
            paste0("WARNING: Multilength output, separated by %%%%%.")
    }

    out
}
```


The following terms will be used in the initial search as exact matches against EuropePMC (acronyms are not included because they return too many hits):

```{r}
terms <- c(
    # "CKM",
    # "CRM",
    "cardiovascular-kidney-metabolic",
    "cardiovascular-renal-metabolic",
    "cardio-kidney-metabolic",
    "cardio-renal-metabolic"
)
```

Notes: 
1. The words "syndrome" & "disease" words are not included to ensure maximal flexibility.
2. "cardiometabolic" and "cardiorenal" were considered for use, but the number of publications using these terms is too great.
3. It appears the wildcard character will not be appropriate for this search. It cannot be used more than once in a word grouping (e.g. "cardio*kidney*metabolic" returns nothing), and while it does work as expected when used singularly, the number of hits is too great to be useful (e.g. "cardio*metabolic" returns numerous hits that have only "cardio-metabolic").

Search will not be limited to Open Access at this time, given how new this disease is. 

_The output will be saved to a file to avoid need of repeat API call (until the next desired updated)._

```{r}
search_str <- terms |>
    DO.utils::sandwich_text('"') |> 
    paste0(collapse = " OR ") #|>
    # DO.utils::sandwich_text(c('OPEN_ACCESS:y AND (', ')'))

if (!file.exists(data_file)) {
    res <- europepmc::epmc_search(search_str, synonym = FALSE, limit = 20000)

    save(res, file = data_file)
} else {
    load(data_file)
}
```

The number of publication hits was `r format(nrow(res), big.mark = ",")`, including `r format(sum(res$isOpenAccess == "Y"), big.mark = ",")` that are Open Access.

The full text of Open Access publications will also be downloaded for this analysis.

```{r}
use_bookid <- "bookid" %in% names(res)

res_tidy <- res |>
    dplyr::filter(
        !stringr::str_detect(pubType, "retract|preprint"),
        !(is.na(pmcid) & is.na(bookid))
    ) |>
    dplyr::select(
        "id", "title", "pubYear", pubDate = "firstPublicationDate",
        "pmcid", if (use_boookid) { "bookid" }) |>
    dplyr::mutate(
        pubDate = lubridate::as_date(pubDate),
        pubYear = lubridate::year(pubDate)
    )

if (!exists("res_sample")) {
    res_sample <- res_tidy |>
        dplyr::slice_sample(prop = 0.25, by = pubYear)

    save(res, res_sample, file = data_file)
}

if (!exists("res_ftxt")) {
    res_ftxt <- res_sample |>
        dplyr::rowwise() |>
        dplyr::mutate(ft_xml = get_ftxt_safely(pmcid, if (use_bookid) { bookid } else { NA})) |>
        dplyr::mutate(ft = parse_ftxt_xml(ft_xml, "//body"))
    
    save(res, res_sample, res_ftxt, file = data_file)
}
```


# Evaluating Usage

Extracting all these values from the full text of the sample publications and all the titles (in a case-insensitive manner).
```{r}
regex_str <- terms |>
    DO.utils::length_sort(decreasing = TRUE) |>
    paste0(collapse = "|")

hsp <- res_tidy |>
    dplyr::left_join(
        res_ftxt,
        by = c("id", "title", "pubYear", "pubDate", "pmcid", "bookid")
    ) |>
    dplyr::select("id", "pubDate", "title", "ft") |>
    dplyr::mutate(
        title_match = stringr::str_extract_all(
            .data$title,
            stringr::regex(regex_str, ignore_case = TRUE)
        ),
        ft_match = stringr::str_extract_all(
            .data$ft,
            stringr::regex(regex_str, ignore_case = TRUE)
        )
    ) |>
    tidyr::unnest(title_match, keep_empty = TRUE) |>
    tidyr::unnest(ft_match, keep_empty = TRUE) |>
    dplyr::mutate(ft = !is.na(ft))
```

The number of publications with and without matches in their titles or full text, noting whether their full-text was obtained are as follows:

```{r}
hsp |>
    dplyr::summarize(
        title_match = any(!is.na(title_match)),
        ft_match = any(!is.na(ft_match)),
        ft = unique(ft),
        .by = "id"
    ) |>
    dplyr::count(ft, title_match, ft_match) |>
    dplyr::mutate(pct = round(n / sum(n) * 100, 2)) |>
    dplyr::rename(ft_obtained = "ft")
```
Most of the publications do not contain one of these terms, including ~20% of those that were downloaded as full text. Any non-matches will just be dropped for the analysis of names.

```{r}
hsp_match <- hsp |>
    tidyr::pivot_longer(
        title_match:ft_match,
        names_to = c("source", ".value"),
        names_sep = "_",
        values_drop_na = TRUE
    ) |>
    dplyr::mutate(match = stringr::str_to_lower(.data$match)) |>
    dplyr::select(-"title") |>
    # only count once per publication and title/full-text
    DO.utils::collapse_col("match")
```

The terms of interest that and whether they were found is as follows:
```{r}
terms_match <- tibble::tibble(
    terms = terms,
    match = stringr::str_to_lower(terms)
)

terms_match |>
    dplyr::left_join(dplyr::count(hsp_match, match), by = "match") |>
    dplyr::mutate(n = tidyr::replace_na(n, 0)) |>
    dplyr::select(-"match") |>
    dplyr::arrange(dplyr::desc(.data$n))
```
Only two terms weren't found. One is likely historical and may not be available in a computable form and the other looks like it may be a misspelling.

The current name in DO is not in the top 5 most common names in the literature.

Organized by publication date and binned into year intervals, the results are as follows:
```{r}
hsp_df <- hsp_match |>
    dplyr::left_join(terms_match, by = "match") |>
    dplyr::select(-"match") |>
    dplyr::rename(term = "terms") |>
    dplyr::filter(!is.na(term))

hsp_colors <- hues::iwanthue(dplyr::n_distinct(hsp_df$term))
    
ggplot2::ggplot(hsp_df) +
    ggplot2::geom_freqpoly(
        ggplot2::aes(x = pubDate, color = term),
        binwidth = 365
    ) +
    ggplot2::scale_color_manual(values = hsp_colors) +
    ggplot2::facet_wrap(~ source, ncol = 1, scales = "free_y")
```

Hmm... the oldest uses are quite a long time ago and make the graph a bit hard to read. Subsetting the graph to after the year 2000:
```{r warning=FALSE}
g <- ggplot2::ggplot(hsp_df) +
    ggplot2::geom_freqpoly(
        ggplot2::aes(x = pubDate, color = term, group = term),
        binwidth = 365,
        linewidth = 1
    ) +
    ggplot2::scale_color_manual(values = hsp_colors) +
    ggplot2::scale_x_date(
        date_breaks = "5 years",
        date_labels = "%Y"
    ) +
    ggplot2::facet_wrap(~ source, ncol = 1, scales = "free_y") +
    ggplot2::coord_cartesian(
        xlim = c(as.Date("2000-01-01"), as.Date("2025-01-01"))
    ) +
    ggplot2::theme_minimal()

plotly::ggplotly(g)
```

Based on this, it seems that for most of the DO's history, this disease has primarily been called "Henoch-Schönlein purpura" which was the original name of the published disease. It's likely this was not used by the DO because of potential encoding issues with the "ö" and, in the literature, it appears that using a normal "o" was next most commmon, but the current label in DO has barely been used and should likely NOT be used as the label.

Furthermore, since ~2020, the name "IgA vasculitis" has been used more frequently in publications. An NIDDK source also suggests that this disease has been renamed to IgA vasculitis: https://www.niddk.nih.gov/health-information/kidney-disease/iga-vasculitis. Although the full name "immunoglobulin A vasculitis" seems like it would be more appropriate, it is used far less frequently. Together, these suggest the primary name should be updated to "IgA vasculitis".

The initialisms for these names are used far less frequently, with HSP being greater than IgAV. It's possible that HSP represents something else given it's more common (e.g. "heat shock proteins", "hereditary spastic paraplegia").


### Confirming relevance of HSP

Just to confirm that HSP is used in conjunction with this disease name:
```{r}
dplyr::count(hsp_df, term, source) |>
    dplyr::filter(term == "HSP")
```

It seems to appear in titles. Looking at which those are:
```{r}
dplyr::filter(hsp, title_match == "HSP")$title
```

Well, it is used to mean this disease but clearly also used for other things. It seems like it should be added as a synonym.


# CONCLUSION

New primary label: "IgA vasculitis"

Ensure all these terms are included as synonyms in the DO:


```{r results='asis', echo = FALSE}
cat(
    sprintf('* %s   ',terms[terms %in% hsp_df$term]),
    sep = '\n'
)
```

Even those these terms were not found in this analysis, there's a possibility that they have been used, historically or perhaps accidentally, so they will be left in the ontology, if they are there already but not added.
```{r results="asis", echo = FALSE}
cat(
    sprintf('* %s   ',terms[!terms %in% hsp_df$term]),
    sep = '\n'
)
```


On review, it appears that only "Autoimmune purpura" is already in the ontology. That will be kept and the other potential synonym will not be added.
